Models including an interaction of treatment with baseline risk were better able to predict absolute benefit

### 2021-07-05 submission/manuscript.rmd
include dvk's comments to discussion

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
 **Objective:** Baseline risk is an important determinant of the absolute effect of a treatment for an individual patient. We aimed to compare different risk-based methods for predicting individualized treatment effects with simulations of the RCT setting. **Study Design** and Setting: We sampled patients from a superpopulation with diverse assumptions for a baseline prognostic index of risk (PI), and for the shape of the interaction between the PI and treatment (no, linear or quadratic interaction). In each sample, we fitted different models for predicting absolute benefit: a model with the PI and a constant relative treatment effect and models including an interaction of treatment assignment with the PI, with 4 quarters of the PI, and with nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots). We also used an adaptive model selection approach based on Akaike’s Information Criterion. We evaluated predictive performance in the superpopulation using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied the sample size,  the strength of the treatment effect, and the discriminative ability of the PI. **Results:** Models including a linear interaction of the PI with treatment had (close to?) optimal performance under most simulation scenarios. More flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model. The adaptive approach performed similarly to the best-performing method in each scenario. **Conclusion:** Under most circumstances, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The constant treatment effect model, despite having adequate performance in the
presence of weak treatment effect heterogeneity on the relative scale, quickly
broke down with stronger deviations from constant relative treatment effects. In
these cases, the stratified approach generally had lower error rates compared to
the constant treatment effect model. In concept, the stratified approach lies
between the constant effects model and smoother approaches, only assuming
constant treatment effects within strata of predicted risk and, therefore, is
more sensitive to treatment effect heterogeneity. Stepwise treatment benefit
estimates are very useful for demonstrating treatment effect heterogeneity –
because estimating treatment effect requires groups of patients rather than
individual patients –but are not helpful for making individualized absolute
benefit predictions.

Increasing the discriminative ability of the risk model – by increasing the
predictor coefficients of the true risk model – reduced RMSE for all methods,
similar to increasing the sample size. This increase in discriminative ability
translates in higher variability of predicted risks, which, in turn, allows the
considered methods to better follow the evolution of treatment benefit. As
expected, the increase in discriminative ability of the risk model also led to
higher values of c-for-benefit. Even though risk model performance is very
important for the ability of risk-based methods to predict treatment benefit,
prediction model development was outside the scope of this work and has already
been studied extensively [@vanKlaveren2019; @Burke2014; @Abadie2018].

@@

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment interaction with
the prognostic performed best in a wide range of scenarios. More flexible models
with restricted cubic splines required larger sample sizes and higher AUC of the
PI to outperform the linear model. An adaptive approach, selecting the model
with optimal AIC, had satisfactory performance.

>>>>

---
title: "Models including an interaction of treatment with baseline risk were better able to predict absolute benefit"
abstract: |
 **Objective:** To compare different risk-based methods for predicting individualized treatment effects using simulations of the RCT setting. **Study Design and Setting:** We sampled patients using diverse assumptions for a baseline prognostic index of risk (PI) and the shape of its intereraction with treatment (none, linear or quadratic). In each sample we predicted absolute benefit using: models with the PI and a constant relative treatment effect; models including an interaction of treatment with the PI; stratification in 4 quarters of the PI; nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive model selection approach based on Akaike’s Information Criterion. We evaluated predictive performance using root mean squared error and measures of discrimination and calibration for benefit. Starting from a base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample size, treatment effect strength, and the PI's discriminative ability. **Results:** Models including a PI by treatment interaction performed better under most simulation settings. Flexible models with restricted cubic splines required larger sample sizes and higher AUC of the PI to outperform the linear model. The adaptive approach performed similarly to the best-performing method in most scenarios. **Conclusion:** Usually, a model with a linear interaction of the PI with treatment is better able to predict absolute treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The constant treatment effect model, despite having adequate performance in the
presence of weak treatment effect heterogeneity on the relative scale, quickly
broke down with stronger deviations from constant relative treatment effects. In
these cases, the stratified approach generally had lower error rates compared to
the constant treatment effect model. Stepwise treatment benefit estimates are
very useful for demonstrating treatment effect heterogeneity--because
estimating treatment effect requires groups of patients rather than individual
patients--but are not helpful for making individualized absolute benefit
predictions.

Increasing the discriminative ability of the risk model – by increasing the
predictor coefficients of the true risk model – reduced RMSE for all methods.
This increase in discriminative ability translates in higher variability of
predicted risks, which, in turn, allows the considered methods to better capture
absolute treatment benefits. As a consequence, the increase in discriminative
ability of the risk model also led to higher values of c-for-benefit. Even
though risk model performance is very important for the ability of risk-based
methods to predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

@@

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment interaction with
the prognostic performed best in a wide range of scenarios. More flexible models
with restricted cubic splines required larger sample sizes and higher AUC of the
PI to outperform the linear model. An adaptive approach, selecting the model
with optimal AIC, had comparable performance to the best performing approach in
most of the scenarios.

### 2021-07-05 submission/manuscript.tex
include dvk's comments to discussion

\title{Continuous risk-based predictive approaches to treatment effect
heterogeneity: A simulation study}

@@

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\begin{abstract}
\textbf{Objective:} Baseline risk is an important determinant of the
absolute effect of a treatment for an individual patient. We aimed to
compare different risk-based methods for predicting individualized
treatment effects with simulations of the RCT setting. \textbf{Study
Design} and Setting: We sampled patients from a superpopulation with
diverse assumptions for a baseline prognostic index of risk (PI), and
for the shape of the interaction between the PI and treatment (no,
linear or quadratic interaction). In each sample, we fitted different
models for predicting absolute benefit: a model with the PI and a
constant relative treatment effect and models including an interaction
of treatment assignment with the PI, with 4 quarters of the PI, and with
nonlinear transformations of the PI (restricted cubic splines with 3, 4
and 5 knots). We also used an adaptive model selection approach based on
Akaike's Information Criterion. We evaluated predictive performance in
the superpopulation using root mean squared error and measures of
discrimination and calibration for benefit. Starting from a base case
scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI
0.75), we varied the sample size, the strength of the treatment effect,
and the discriminative ability of the PI. \textbf{Results:} Models
including a linear interaction of the PI with treatment had (close to?)
optimal performance under most simulation scenarios. More flexible
models with restricted cubic splines required larger sample sizes and
higher AUC of the PI to outperform the linear model. The adaptive
approach performed similarly to the best-performing method in each
scenario. \textbf{Conclusion:} Under most circumstances, a model with a
linear interaction of the PI with treatment is better able to predict
absolute treatment benefit.
\end{abstract}

@@

The constant treatment effect model, despite having adequate performance
in the presence of weak treatment effect heterogeneity on the relative
scale, quickly broke down with stronger deviations from constant
relative treatment effects. In these cases, the stratified approach
generally had lower error rates compared to the constant treatment
effect model. In concept, the stratified approach lies between the
constant effects model and smoother approaches, only assuming constant
treatment effects within strata of predicted risk and, therefore, is
more sensitive to treatment effect heterogeneity. Stepwise treatment
benefit estimates are very useful for demonstrating treatment effect
heterogeneity -- because estimating treatment effect requires groups of
patients rather than individual patients --but are not helpful for
making individualized absolute benefit predictions.

Increasing the discriminative ability of the risk model -- by increasing
the predictor coefficients of the true risk model -- reduced RMSE for
all methods, similar to increasing the sample size. This increase in
discriminative ability translates in higher variability of predicted
risks, which, in turn, allows the considered methods to better follow
the evolution of treatment benefit. As expected, the increase in
discriminative ability of the risk model also led to higher values of
c-for-benefit. Even though risk model performance is very important for
the ability of risk-based methods to predict treatment benefit,
prediction model development was outside the scope of this work and has
already been studied extensively {[}5,12,13{]}.

>>>>

\title{Models including an interaction of treatment with baseline risk
were better able to predict absolute benefit}

@@

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods for
predicting individualized treatment effects using simulations of the RCT
setting. \textbf{Study Design and Setting:} We sampled patients using
diverse assumptions for a baseline prognostic index of risk (PI) and the
shape of its intereraction with treatment (none, linear or quadratic).
In each sample we predicted absolute benefit using: models with the PI
and a constant relative treatment effect; models including an
interaction of treatment with the PI; stratification in 4 quarters of
the PI; nonlinear transformations of the PI (restricted cubic splines
with 3, 4 and 5 knots); an adaptive model selection approach based on
Akaike's Information Criterion. We evaluated predictive performance
using root mean squared error and measures of discrimination and
calibration for benefit. Starting from a base case scenario (sample size
4,250, treatment odds ratio 0.8, AUC of the PI 0.75), we varied sample
size, treatment effect strength, and the PI's discriminative ability.
\textbf{Results:} Models including a PI by treatment interaction
performed better under most simulation settings. Flexible models with
restricted cubic splines required larger sample sizes and higher AUC of
the PI to outperform the linear model. The adaptive approach performed
similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Usually, a model with a linear interaction of the
PI with treatment is better able to predict absolute treatment benefit.
\end{abstract}

@@

The constant treatment effect model, despite having adequate performance
in the presence of weak treatment effect heterogeneity on the relative
scale, quickly broke down with stronger deviations from constant
relative treatment effects. In these cases, the stratified approach
generally had lower error rates compared to the constant treatment
effect model. Stepwise treatment benefit estimates are very useful for
demonstrating treatment effect heterogeneity--because estimating
treatment effect requires groups of patients rather than individual
patients--but are not helpful for making individualized absolute benefit
predictions.

Increasing the discriminative ability of the risk model -- by increasing
the predictor coefficients of the true risk model -- reduced RMSE for
all methods. This increase in discriminative ability translates in
higher variability of predicted risks, which, in turn, allows the
considered methods to better capture absolute treatment benefits. As a
consequence, the increase in discriminative ability of the risk model
also led to higher values of c-for-benefit. Even though risk model
performance is very important for the ability of risk-based methods to
predict treatment benefit, prediction model development was outside the
scope of this work and has already been studied extensively
{[}5,12,13{]}.
